### 2025-07-29 paper/quarto_aaai/paper.tex

CEs have also been used to improve models in the natural language
processing domain. For example, \citet{wu2021polyjuice} propose
\emph{Polyjuice}, a general-purpose CE generator for language models and
demonstrate that the augmentation of training data with \emph{Polyjuice}
improves robustness in a number of tasks, while
\citet{luu2023counterfactual} introduce the \emph{Counterfactual
Adversarial Training} (CAT) framework that aims to improve
generalization and robustness of language models by generating
counterfactuals for training samples that are subject to high predictive
uncertainty.

@@

The characterization of faithfulness and plausibility in Def.
\ref{def-explainability} follows \citet{altmeyer2024faithful}, with
adapted notation. Intuitively, plausible counterfactuals are consistent
with the data and faithful counterfactuals are consistent with what the
model has learned about the input data. Ac tionability constraints in
Def. \ref{def-explainability} vary and depend on the context in which
\(\mathbf{M}_\theta\) is deployed. In this work, we choose to only
consider domain and mutability constraints for individual features
\(x_d\) for \(d=1,...,D\). We also limit ourselves to classification
tasks for reasons discussed in Section~\ref{sec-discussion}.

Let \(\mathbf{x}_t^\prime\) for \(t=0,...,T\) denote a counterfactual
generated through gradient descent over \(T\) iterations as originally
proposed by \citet{wachter2017counterfactual}. CT adopts gradient-based
CE search in training to generate on-the-fly model explanations
\(\mathbf{x}^\prime\) for the training samples. We use the term
\emph{nascent} to denote interim counterfactuals
\(\mathbf{x}_{t\leq T}^\prime\) that have not yet converged. As we
explain below, these nascent counterfactuals can be stored and
repurposed as adversarial examples. Conversely, we consider
counterfactuals \(\mathbf{x}_T^\prime\) as \emph{mature} explanations if
they have either exhausted all \(T\) iterations or converged by reaching
a pre-specified threshold, \(\tau\), for the predicted probability of
the target class:
\(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\),
where \(\mathcal{S}\) is the softmax function.

Formally, we propose the following counterfactual training objective to
train explainable (as in Def. \ref{def-explainability}) models,
\begin{equation}\phantomsection\label{eq-obj}{
\begin{aligned}
&\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) \\+ &\lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) + \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
\end{aligned}
}\end{equation} where \(\text{yloss}(\cdot)\) is any classification loss
that induces discriminative performance (e.g., cross-entropy). The
second and third terms are explained in detail below. For now, they can
be summarized as inducing explainability directly and indirectly by
penalizing (1) the contrastive divergence, \(\text{div}(\cdot)\),
between mature counterfactuals \(\mathbf{x}_T^\prime\) and observed
samples \(\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}\) in the
target class \(y^+\), and (2) the adversarial loss,
\(\text{advloss}(.)\), wrt. nascent counterfactuals
\(\mathbf{x}_{t\leq T}^\prime\). Finally, \(\text{ridge}(\cdot)\)
denotes a Ridge penalty (\(\ell_2\)-norm) that regularizes the magnitude
of the energy terms involved in \(\text{div}(\cdot)\)
\citep{du2019implicit}. The trade-offs between these components are
adjusted through \(\lambda_{\text{div}}\), \(\lambda_{\text{adv}}\) and
\(\lambda_{\text{reg}}\). The full training regime is sketched out in
Algorithm \ref{alg-experiment}.

@@

To generate samples, \citet{grathwohl2020your} use Stochastic Gradient
Langevin Dynamics (SGLD) with an uninformative prior for initialization
but we depart from their methodology: we propose to leverage
counterfactual explainers to generate counterfactuals of observed
training samples. Specifically, we have:
\begin{equation}\phantomsection\label{eq-div}{
\text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) = \mathcal{E}_\theta(\mathbf{x}^+,y) - \mathcal{E}_\theta(\mathbf{x}_T^\prime,y)
}\end{equation} where \(\mathcal{E}_\theta(\cdot)\) denotes the energy
function defined as
\(\mathcal{E}_\theta(\mathbf{x},y)=-\mathbf{M}_\theta(\mathbf{x})[y^+]\),
with \(y^+\) denoting the index of the randomly drawn target class,
\(y^+ \sim p(y)\). Conditional on the target class \(y^+\),
\(\mathbf{x}_T^\prime\) denotes a mature counterfactual for a randomly
sampled factual from a non-target class generated with a gradient-based
CE generator for up to \(T\) iterations. Intuitively, the gradient of
Equation~\ref{eq-div} decreases the energy of observed training samples
(positive samples) while increasing the energy of counterfactuals
(negative samples) \citep{du2019implicit}. As the counterfactuals get
more plausible (Def. \ref{def-explainability}) during training, these
opposing effects gradually balance each other out
\citep{lippe2024uvadlc}.

@@

Based on our analysis in Section~\ref{sec-lit}, counterfactuals
\(\mathbf{x}^\prime\) can be repurposed as additional training samples
\citep{balashankar2023improving, luu2023counterfactual} or adversarial
examples \citep{freiesleben2022intriguing, pawelczyk2022exploring}. This
leaves some flexibility with regards to the choice for the
\(\text{advloss}(\cdot)\) term in Equation~\ref{eq-obj}. An intuitive
functional form, but likely not the only sensible choice, is inspired by
adversarial training: \begin{equation}\phantomsection\label{eq-adv}{
\begin{aligned}
\text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y};\varepsilon)&=\text{yloss}(\mathbf{M}_\theta(\mathbf{x}_{t_\varepsilon}^\prime),\mathbf{y}) \\
t_\varepsilon &= \max_t \{t: ||\Delta_t||_\infty < \varepsilon\}
\end{aligned}
}\end{equation} Under this choice, we consider nascent counterfactuals
\(\mathbf{x}_{t\leq T}^\prime\) as AEs as long as the magnitude of the
perturbation to any single feature is at most \(\varepsilon\). This is
closely aligned with \citet{szegedy2013intriguing} who define an
adversarial attack as an ``imperceptible non-random perturbation''.
Thus, we work with a different distinction between CE and AE than
\citet{freiesleben2022intriguing} who considers misclassification as the
distinguishing feature of adversarial examples. One of the key
observations of this work is that we can leverage CEs during training
and get AEs essentially for free to reap the aforementioned benefits of
adversarial training.

@@

We make use of nine classification datasets common in the CE/AR
literature. Four of them are synthetic with two classes and different
characteristics: linearly separable clusters (\emph{LS}), overlapping
clusters (\emph{OL}), concentric circles (\emph{Circ}), and interlocking
moons (\emph{Moon}). Next, we have four real-world binary tabular
datasets: \emph{Adult} (Census data) of \citet{becker1996adult2},
California housing (\emph{CH}) of \citet{pace1997sparse}, Default of
Credit Card Clients (\emph{Cred}) of \citet{yeh2016default}, and Give Me
Some Credit (\emph{GMSC}) from \citet{kaggle2011give}. Finally, for the
convenience of illustration, we use the 10-class \emph{MNIST}
\citep{lecun1998mnist}.

>>>>

CEs have also been used to improve models in the natural language
processing domain. For example, \citet{wu2021polyjuice} propose
\emph{Polyjuice}, a general-purpose CE generator for language models,
and demonstrate that the augmentation of training data with
\emph{Polyjuice} improves robustness in a number of tasks.
\citet{luu2023counterfactual} introduce the \emph{Counterfactual
Adversarial Training} (CAT) framework that aims to improve
generalization and robustness of language models by generating
counterfactuals for training samples that are subject to high predictive
uncertainty.

@@

The characterization of faithfulness and plausibility in Def.
\ref{def-explainability} follows \citet{altmeyer2024faithful}, with
adapted notation. Intuitively, plausible counterfactuals are consistent
with the data and faithful counterfactuals are consistent with what the
model has learned about the input data. Actionability constraints in
Def. \ref{def-explainability} vary and depend on the context in which
\(\mathbf{M}_\theta\) is deployed. In this work, we choose to only
consider domain and mutability constraints for individual features
\(x_d\) for \(d=1,...,D\). We also limit ourselves to classification
tasks for reasons discussed in Section~\ref{sec-discussion}.

Let \(\mathbf{x}_t^\prime\) for \(t=0,...,T\) denote a counterfactual
generated through gradient descent over \(T\) iterations as originally
proposed by \citet{wachter2017counterfactual}. CT adopts gradient-based
CE search in training to generate on-the-fly model explanations
\(\mathbf{x}^\prime\) for the training samples. We use the term
\emph{nascent} to denote interim counterfactuals
\(\mathbf{x}_{\text{AE}}^\prime\) that have not yet converged. As we
explain below, these nascent counterfactuals can be stored and
repurposed as adversarial examples. Conversely, we consider
counterfactuals \(\mathbf{x}_{\text{CE}}^\prime\) as \emph{mature}
explanations if they have converged within the \(T\) iterations by
reaching a pre-specified threshold, \(\tau\), for the predicted
probability of the target class:
\(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\),
where \(\mathcal{S}\) is the softmax function.

Formally, we propose the following counterfactual training objective to
train explainable (as in Def. \ref{def-explainability}) models,
\begin{equation}\phantomsection\label{eq-obj}{
\begin{aligned}
&\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_{\text{CE}}^\prime,y;\theta) \\+ &\lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{\text{AE}}^\prime),\mathbf{y}) + \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_{\text{CE}}^\prime,y;\theta)
\end{aligned}
}\end{equation} where \(\text{yloss}(\cdot)\) is any classification loss
that induces discriminative performance (e.g., cross-entropy). The
second and third terms are explained in detail below. For now, they can
be summarized as inducing explainability directly and indirectly by
penalizing (1) the contrastive divergence, \(\text{div}(\cdot)\),
between mature counterfactuals \(\mathbf{x}_{\text{CE}}^\prime\) and
observed samples \(\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}\)
in the target class \(y^+\), and (2) the adversarial loss,
\(\text{advloss}(.)\), wrt. nascent counterfactuals
\(\mathbf{x}_{\text{AE}}^\prime\). Finally, \(\text{ridge}(\cdot)\)
denotes a Ridge penalty (\(\ell_2\)-norm) that regularizes the magnitude
of the energy terms involved in \(\text{div}(\cdot)\)
\citep{du2019implicit}. The trade-offs between these components are
adjusted through \(\lambda_{\text{div}}\), \(\lambda_{\text{adv}}\) and
\(\lambda_{\text{reg}}\). The full training regime is sketched out in
Algorithm \ref{alg-experiment}.

@@

To generate samples, \citet{grathwohl2020your} use Stochastic Gradient
Langevin Dynamics (SGLD) with an uninformative prior for initialization
but we depart from their methodology: we propose to leverage
counterfactual explainers to generate counterfactuals of observed
training samples. Specifically, we have:
\begin{equation}\phantomsection\label{eq-div}{
\text{div}(\mathbf{x}^+,\mathbf{x}_{\text{CE}}^\prime,y;\theta) = \mathcal{E}_\theta(\mathbf{x}^+,y) - \mathcal{E}_\theta(\mathbf{x}_{\text{CE}}^\prime,y)
}\end{equation} where \(\mathcal{E}_\theta(\cdot)\) denotes the energy
function defined as
\(\mathcal{E}_\theta(\mathbf{x},y)=-\mathbf{M}_\theta(\mathbf{x})[y^+]\),
with \(y^+\) denoting the index of the randomly drawn target class,
\(y^+ \sim p(y)\). Conditional on the target class \(y^+\),
\(\mathbf{x}_{\text{CE}}^\prime\) denotes a mature counterfactual for a
randomly sampled factual from a non-target class generated with a
gradient-based CE generator for up to \(T\) iterations. Intuitively, the
gradient of Equation~\ref{eq-div} decreases the energy of observed
training samples (positive samples) while increasing the energy of
counterfactuals (negative samples) \citep{du2019implicit}. As the
counterfactuals get more plausible (Def. \ref{def-explainability})
during training, these opposing effects gradually balance each other out
\citep{lippe2024uvadlc}.

@@

Based on our analysis in Section~\ref{sec-lit}, counterfactuals
\(\mathbf{x}^\prime\) can be repurposed as additional training samples
\citep{balashankar2023improving, luu2023counterfactual} or adversarial
examples \citep{freiesleben2022intriguing, pawelczyk2022exploring}. This
leaves some flexibility with regards to the choice for the
\(\text{advloss}(\cdot)\) term in Equation~\ref{eq-obj}. An intuitive
functional form, but likely not the only sensible choice, is inspired by
adversarial training: \begin{equation}\phantomsection\label{eq-adv}{
\begin{aligned}
\text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{\text{AE}}^\prime),\mathbf{y};\varepsilon)&=\text{yloss}(\mathbf{M}_\theta(\mathbf{x}_{t_\varepsilon}^\prime),\mathbf{y}) \\
t_\varepsilon &= \max_t \{t: ||\Delta_t||_\infty < \varepsilon\}
\end{aligned}
}\end{equation} Under this choice, we consider nascent counterfactuals
\(\mathbf{x}_{\text{AE}}^\prime\) as AEs as long as the magnitude of the
perturbation to any single feature is at most \(\varepsilon\). This is
closely aligned with \citet{szegedy2013intriguing} who define an
adversarial attack as an ``imperceptible non-random perturbation''.
Thus, we work with a different distinction between CE and AE than
\citet{freiesleben2022intriguing} who considers misclassification as the
distinguishing feature of adversarial examples. One of the key
observations of this work is that we can leverage CEs during training
and get AEs essentially for free to reap the aforementioned benefits of
adversarial training.

@@

We make use of nine classification datasets common in the CE/AR
literature. Four of them are synthetic with two classes and different
characteristics: linearly separable Gaussian clusters (\emph{LS}),
overlapping clusters (\emph{OL}), concentric circles (\emph{Circ}), and
interlocking moons (\emph{Moon}). Next, we have four real-world binary
tabular datasets: \emph{Adult} (Census data) of
\citet{becker1996adult2}, California housing (\emph{CH}) of
\citet{pace1997sparse}, Default of Credit Card Clients (\emph{Cred}) of
\citet{yeh2016default}, and Give Me Some Credit (\emph{GMSC}) from
\citet{kaggle2011give}. Finally, for the convenience of illustration, we
use the 10-class \emph{MNIST} \citep{lecun1998mnist}.

### 2025-07-29 paper/sections/experiments.qmd

We make use of nine classification datasets common in the CE/AR literature. Four of them are synthetic with two classes and different characteristics: linearly separable clusters (*LS*), overlapping clusters (*OL*), concentric circles (*Circ*), and interlocking moons (*Moon*). Next, we have four real-world binary tabular datasets: *Adult* (Census data) of @becker1996adult2, California housing (*CH*) of @pace1997sparse, Default of Credit Card Clients (*Cred*) of @yeh2016default, and Give Me Some Credit (*GMSC*) from @kaggle2011give. Finally, for the convenience of illustration, we use the 10-class *MNIST* [@lecun1998mnist].

@@

Test accuracies on adversarially perturbed data are shown in @fig-acc. The perturbations size, $\varepsilon\in[0,0.1]$, increases along the horizontal axis and includes zero, corresponding to standard test accuracy for non-perturbed data. For all synthetic datasets, predictive performance of CT is virtually identical to the baseline and unaffected by perturbations. For all real-world datasets, we find that CT substantially improves robustness: while in some cases baseline accuracies drop to essentially zero for large enough perturbation sizes, accuracies of CT models remain remarkably robust.

@@

Secondly, the strength of the energy regularization, $\lambda_{\text{reg}}$ is highly impactful and should be set sufficiently high to avoid common problems associated with exploding gradients. The sensitivity with respect to $\lambda_{\text{div}}$ and $\lambda_{\text{adv}}$ is much less evident. While high values of $\lambda_{\text{reg}}$ may increase the variability in outcomes when combined with high values of $\lambda_{\text{div}}$ or $\lambda_{\text{adv}}$, this effect is not particularly pronounced.

>>>>

We make use of nine classification datasets common in the CE/AR literature. Four of them are synthetic with two classes and different characteristics: linearly separable Gaussian clusters (*LS*), overlapping clusters (*OL*), concentric circles (*Circ*), and interlocking moons (*Moon*). Next, we have four real-world binary tabular datasets: *Adult* (Census data) of @becker1996adult2, California housing (*CH*) of @pace1997sparse, Default of Credit Card Clients (*Cred*) of @yeh2016default, and Give Me Some Credit (*GMSC*) from @kaggle2011give. Finally, for the convenience of illustration, we use the 10-class *MNIST* [@lecun1998mnist].

@@

Test accuracies on adversarially perturbed data are shown in @fig-acc. The perturbation size, $\varepsilon\in[0,0.1]$, increases along the horizontal axis and includes zero, corresponding to standard test accuracy for non-perturbed data. For all synthetic datasets, predictive performance of CT is virtually identical to the baseline and unaffected by perturbations. For all real-world datasets, we find that CT substantially improves robustness: while in some cases baseline accuracies drop to essentially zero for large enough perturbation sizes, accuracies of CT models remain remarkably robust.

@@

Secondly, the strength of the energy regularization, $\lambda_{\text{reg}}$, is highly impactful and should be set sufficiently high to avoid common problems associated with exploding gradients. The sensitivity with respect to $\lambda_{\text{div}}$ and $\lambda_{\text{adv}}$ is much less evident. While high values of $\lambda_{\text{reg}}$ may increase the variability in outcomes when combined with high values of $\lambda_{\text{div}}$ or $\lambda_{\text{adv}}$, this effect is not particularly pronounced.

### 2025-07-29 paper/sections/other/notation.qmd

- $\mathcal{Y}$: The output domain.
- $y^+$: The target class and also the index of the target class.
- $y^-$: The non-target class and also the index of non-the target class.
- $\mathcal{X}$: The input domain.
- $\mathbf{x}$: a single training sample.
- $\mathbf{x}^\prime$: a counterfactual.
- $t=1,...,T$: Step indicator for counterfactual search iterations.
- $\mathbf{x}^\prime_{t \leq T}$: a nascent counterfactual, defined as a counterfactual that has not yet converged.
- $\mathbf{x}^\prime_{T}$: a mature counterfactual, defined as a counterfactual that has either passed a threshold probability or exhausted all $T$ steps.
- $\mathbf{x}^+$: a training sample in the target class (ground-truth).
- $\mathbf{y}^+$: The one-hot encoded output vector for the target class.
- $\theta$: Model parameters (unspecified).
- $\Theta$: Matrix of parameters. 
- $\mathbf{M}(\cdot)$: linear predictions (logits) of the classifier.

>>>>

- $\mathcal{Y}$: The output domain.
- $y^+$: The target class and also the index of the target class.
- $y^-$: The non-target class and also the index of non-the target class.
- $\mathcal{X}$: The input domain.
- $\mathbf{x}$: a single training sample.
- $\mathbf{x}^\prime$: a counterfactual.
- $t=1,...,T$: Step indicator for counterfactual search iterations.
- $\mathbf{x}_{\text{AE}}^\prime$: a nascent counterfactual, defined as a counterfactual that has not yet converged.
- $\mathbf{x}_{\text{CE}}^\prime$: a mature counterfactual, defined as a counterfactual that has either passed a threshold probability or exhausted all $T$ steps.
- $\mathbf{x}^+$: a training sample in the target class (ground-truth).
- $\mathbf{y}^+$: The one-hot encoded output vector for the target class.
- $\theta$: Model parameters (unspecified).
- $\Theta$: Matrix of parameters. 
- $\mathbf{M}(\cdot)$: linear predictions (logits) of the classifier.
